NMRlipids IV: Headgroup \& glycerol backbone structures, and cation binding in bilayers with PS lipids

### 2018-03-06 Manuscript/manuscriptPS.tex
Update manuscript

\subsection{Headgroup structure in PS and PC mixtures}
Lipid mixtures containing PC and PS lipids are present in biological
systems. Also the Ca2+ binding affinity to bilayers containing
PS lipids is measured from mixtures with PC, because pure PS
bilayers exhibit instant transition to the ordered phase with the
addition of the ions. Therefore, the characterization of mutual interactions
between PC and PS lipid headgroups is necessary in order to
understand PS lipids in biological environments.

Fig. \ref{HGorderparametersPCvsPS} shows the headgroup order parameters
of POPC and POPS lipids in mixtures with various mole fractions from 
simulations and experiments \cite{scherer87,roux90}.
The headgroup order parameters of POPC increase with the increasing
amount of negatively charged POPS, as expected from the 
the electrometer concept \cite{seelig87}. Notably, the concept
seems to be generally valid, at least qualitatively, also for 
other lipid mixed with PC lipids; the PC lipid headgroup order
parameters increase when mixed with negatively charged 
lipids (PS, PI, CL, PA and PG) and remains almost unchaged when 
mixed with neutral lipids (PE and SM) \cite{scherer87} or cholesterol \cite{ferreira13}.
This is, however, not reproduced in the simulation data 
in Fig. \ref{HGorderparametersPCvsPS}. The order parameter for
$\beta$-carbon of POPC decreases and almost negligible increase is 
seen in $\alpha$ order parameter with the addition of PS lipid in CHARMM36 simulation.
This may indicate incorrect interactions between PC and PS headgroups
in simulations. On the other hand, the previously reported overestimated
ion binding affinity may also play a role, as the amount of counterions
is increasing with PS.
\todo{Data from other force fields to be collected before discussion can be finalized.}

@@

@@

The headgroup order parameter changes of POPC and POPS from POPC:POPS (5:1) bilayers
as a function of Ca$2+$ concentration from different
simulations and experiments \cite{roux90} are shown in Fig. \ref{changesWITHCaClPG}.
As shown previously, the changes of PC headgroup order parameters can be
used to measure the amount of bound ions to the bilayer according to the
electrometer concept \cite{akutsu81,seelig87,catte16}.
This would suggest that the Ca2+ binding is overestimated in
MacRog simulation and underestimated in CHARMM36 simulation.
It should be however noted, that MacRog simulation with low
concentrations gives a good agreemet with experiments.
The result from CHARMM simulation is somewhat surprising because
previous study suggested that Ca2+ ions overbind to pure PC lipid
bilayers \cite{catte16}.
\todo{The role of counterions and equilibration should be somehow checked
  and discussed.}
The density profiles of Ca2+ from the simulations are shown in Fig. \ref{CAdensPCPSmixture}.
Overall, it seems that calcium clearly overbinds in MacRog simulations
as expected based on previous results \cite{catte16}.
However, the calcium binding in CHARMM36 simulations seems to
surprisingly low. Thi
\begin{figure*}[]
  \centering
  \includegraphics[width=18cm]{../Figs/CHANGESwithCaClPS.eps}
  \caption{\label{changesWITHCaClPG}
    POPC (left) and POPS (right) headgroup order parameters in lipid bilayer with POPC:POPS (5:1) mixture
    as a function CaCl$_2$ concentration from experiments \cite{roux90} and different simulations.
  }
\end{figure*}
\begin{figure}[]
  \centering
  \includegraphics[width=9cm]{../Figs/CAdensPCPSmixture.eps}
  \caption{\label{CAdensPCPSmixture}
    Ca2+ density profiles from simulations.
  }
  \todo{Ca2+ binding in CHARMM is lower than to neutral membrane. The reason should be found out (probably equilibration).}
  \todo{These are now mass densities, numbers would be probably better.}
  \todo{Counterions should be also included.}
  \todo{Not all the data from MacRog is included.}
\end{figure}

Also dependence of $\beta$-carbon of PG on CaCl$_2$ concentration is compared with
experiments \cite{borle85} in Fig. \ref{PSPGchangesWITHCaCl}. Absolute value of
the order parameter is too large without ions, but rapid decrease due to addition of
CaCl$_2$ is observed in agreement with experiments for systems with 1:1 mixture of POPC and POPG.
In addition, absolute value in systems with CaCl$_2$ is in agreement with experiments.
However, system with 4:1 mixture of POPC and POPG behaves differently, but experimental
data is not available for comparison for this mixture.

\todo{More simulation data for systems with PS lipids and CaCl$_2$ would be probably useful}

%\subsection{Effect of Ca$2+$ binding to PS headgroup}

Also the experimental order parameters of POPS headgroup 
as a function of CaCl$_2$ concentration are shown in Fig. \ref{PSPGchangesWITHCaCl}.
The order parameter changes with added CaCl$_2$ are overestimated in
both simulation models. However, the changes can be considered qualitatively
correct for both segments in CHARMM36 model and in $\beta$ segment in MacRog model. 
\todo{These should be compared to simulations for potential structural interpretation of the changes.}
%\begin{figure}[]
%  \centering
%  \includegraphics[width=9.0cm]{../Figs/PSPGwithCaCl.eps}
%  \caption{\label{PSPGchangesWITHCaCl}
%    PG and PS order parameters a function CaCl$_2$ concentration taken from \cite{borle85} and \cite{roux90}, respectively.
%  }
%  \todo{Get the small concentration data from the inserts}
%\end{figure}

>>>>

\subsection{Headgroup structure in PS and PC mixtures}
Cellular membranes often compose of mixtures of zwitterionic and negatively
charged lipids. The response of PC headgroup to mixtures of differently charged
lipids are collected from different experiments in Fig. \ref{HGorderparametersPCvsPEPSPGchol}.
As expected from the electrometer concept \cite{seelig87}, the headgroup 
order parameters increase with the addition of negatively charged PS and PG lipids,
decrease when mixed with positively charged surfactants and
are less affected by the addition of zwitterionic PE lipids or cholesterol.
In addition to the results summarized in Fig. \ref{HGorderparametersPCvsPEPSPGchol},
also mixtures of PC with negatively charged PI, CL, PA, and zwitterionic SM
follow the eletrometer concept \cite{scherer87}.

%Also the Ca2+ binding affinity to bilayers containing
%PS lipids is measured from mixtures with PC, because pure PS
%bilayers exhibit instant transition to the ordered phase with the
%addition of the ions.
%Therefore, the characterization of mutual interactions
%between PC and PS lipid headgroups is necessary in order to
%understand PS lipids in biological environments.

Fig. \ref{HGorderparametersPCvsPS} shows the PC and PS headgroup order parameters
from POPC:POPS mixtures with various mole fractions from 
simulations and experiments \cite{scherer87,roux90}.
The experimentally observed increase of PC headgroup order parameters with the increasing
amount of negatively charged PS lipid is reproduced in
MacRog simulations with potassium counterions, but not in CHARMM36 simulations
with potassium or sodium conterions. The results can be
understood from the counterion density distributions shown in Fig. \ref{CIdensPSOCmixt}.
Both counterions are more strongly bound to bilayers in
CHARMM36 simulations than potassium in MacRog simulation.
The cation binding probably neutralizes the effect of negatively
charged lipids in CHARMM36 simulations, while the
weaker binding of potassium in MacRog simulations enables the
increase of order parameters with the added amount of PS.

%$\beta$-carbon of POPC decreases and almost negligible increase is 
%seen in $\alpha$ order parameter with the addition of PS lipid 
%This may indicate incorrect interactions between PC and PS headgroups
%in simulations. On the other hand, the previously reported overestimated
%ion binding affinity may also play a role, as the amount of counterions
%is increasing with PS.
%\todo{Data from other force fields to be collected before discussion can be finalized.}

@@

\begin{figure}[]
  \centering
  \includegraphics[width=8.0cm]{../Figs/CIdensPSOCmixt.eps}
  \caption{  Counterion density distributions from PC:PS mixtures.
\label{CIdensPSOCmixt}
  }
\end{figure}

@@

The headgroup order parameter changes of POPC and POPS from POPC:POPS (5:1) bilayers
as a function of Ca$2+$ concentration from different
simulations and experiments \cite{roux90} are shown in Fig. \ref{changesWITHCaClPS}.
The results suggest that Ca2+ ions clearly overbind in simulations with
MacRog model, as expected from previous results for PC lipid bilayers \cite{catte16}.
It should be noted, however, that the lowest concentration (100mM) gives
a good agreemet with experiments \todo{Should be analyze/discuss this further?}.
Surprisingly, the calcium binding seems to be too weak CHARMM36 simulations
\todo{This is most likely due to the new interactions added to CHARMM-GUI:
  http://nmrlipids.blogspot.com/2017/12/nmrlipids-iv-current-status-and.html?showComment=1519913688317\#c5312059274016196300
}.
The density profiles of Ca2+ from the simulations are shown in Fig. \ref{CAdensPCPSmixture}.
\todo{More simulation data for systems with PS lipids and CaCl$_2$ would be probably useful}

\begin{figure*}[]
  \centering
  \includegraphics[width=18cm]{../Figs/CHANGESwithCaClPS.eps}
  \caption{\label{changesWITHCaClPS}
    POPC (left) and POPS (right) headgroup order parameters in lipid bilayer with POPC:POPS (5:1) mixture
    as a function CaCl$_2$ concentration from experiments \cite{roux90} and different simulations.
  }
\end{figure*}
\begin{figure}[]
  \centering
  \includegraphics[width=9cm]{../Figs/CAdensPCPSmixture.eps}
  \caption{\label{CAdensPCPSmixture}
    Ca2+ density profiles from simulations.
  }
  \todo{Ca2+ binding in CHARMM is lower than to neutral membrane. The reason should be found out (probably equilibration).}
  \todo{These are now mass densities, numbers would be probably better.}
  \todo{Counterions should be also included.}
  \todo{Not all the data from MacRog is included.}
\end{figure}

Also the order parameters of PS headgroup as a function of CaCl$_2$ concentration
%from mixtures with PC  
are shown in Fig. \ref{changesWITHCaClPS}. The order parameters exhibit
a strong dependence small CaCl$_2$ concentrations with a rapid saturation. 
The changes of PS headgroup order parameters with added CaCl$_2$ are overestimated in
both simulation models. The response of $\beta$-carbon order parameter could be
interpreted qualitatively correct in simulations, but the general agreement with
experiments is not sufficient to use simulations to interpret the response of
PS headgroup to the added CaCl$_2$.
